\(Q\)-Boson model and relations with integrable hierarchies

### 2024-04-30 draft.tex
improve draft

\address{\noindent 
Instituto de Física Teórica, UNESP-Universidade Estadual Paulista,
R. Dr. Bento T. Ferraz 271, Bl. II, Sao Paulo 01140-070, SP, Brazil\\
\&\\
Instituto de Física, Universidade de S\~ao Paulo,
Rua do Matão Travessa 1371, 05508-090 São Paulo, SP. Brazil
}

@@

\begin{abstract}
This work delves into the complex interplay between the q-boson model,
a quantum integrable system, and classical integrable systems like the
Toda and KP hierarchies. By analyzing scalar products of Bethe states
and correlation functions, we investigate their connections to tau
functions and explore Schur polynomial expansions. This exploration
unravels the intricate relationship between quantum and classical
integrable systems.
\end{abstract}

%\date{\today}
%\keywords{Integrability, Bethe, Schur}
\subjclass[2020]{82B20, 82B23}
\maketitle

\setcounter{tocdepth}{2}
\tableofcontents

@@

Here, we examine the emergence of classical integrable structures
within correlation functions of the q-boson system. This quantum
integrable system characterizes strongly correlated bosons confined to
a one-dimensional chain~\cite{Bogoliubov:1992, Bogoliubov:1997soj,
  Bogoliubov2005}. Notably, this model bears close relation to the AL
(Ablowitz-Ladik) model, which is an integrable discretization of the
nonlinear Schrödinger equation.

@@

Foda and Wheeler~\cite{Foda:2008hn, Wheeler:2010vmq} utilize a rich
set of dualities between the ring of symmetric functions and
\(q\)-deformations of the Heisenberg algebra~\cite{Jing1991, Jing1995}
to explore the combinatorial properties of this spin chain. As a
consequence, it has been shown that the scalar product of off-shell
Bethe states are restricted tau functions of the KP
(Kadomtsev-Petviashvili) hierarchy.

@@

This paper builds upon these developments and aims to extend some of
these results. Firstly, we delve deeper into the connection between
the scalar product and tau functions, exploring them in a wider
context, particularly in the case of the Toda
hierarchy~\cite{Takasaki:2018wsv}. It is important to note that both
the AL and the KP hierarchies are reductions of the Toda
hierarchy. This paper takes steps towards resolving this problem,
which is instrumental in understanding the finer details of these
relationships.

Additionally, we explore other correlation functions within this model
and examine how they fit into the classical integrable context. It is
noteworthy that even when the object is not a tau function itself, it
may have an interesting expansion in terms of different classes of
Schur polynomials, highlighting some of its combinatorial
properties\footnote{Numerous results involving Schur and
Hall-Littlewood polynomials have been confirmed using the
software~\cite{Araujo:2024piv}.}. We also investigate some of these
expansions in this paper.

In Section 2, we provide a review of the phase model and
q-bosons. This section serves to establish notation and emphasize the
key aspects relevant to our analysis.  Section 3 delves into the
analysis of the Phase Model and its connections with the KP and Toda
integrable hierarchies. We explore the general expansion of a phase
model correlation function that yields a Toda system solution.  In
Section 4, we shift our focus to the q-bosons. We discuss a
determinantal formula for the scalar product of two off-shell Bethe
states and its expansion in terms of big, supersymmetric Schur
functions, as well as in terms of Kostka-Foulkes polynomials.
Finally, in Section 5, we conclude with an overview of open problems.

@@

This section introduces the \emph{phase} and \emph{\(q\)-boson}
models~\cite{Bogoliubov:1992, Bogoliubov:1997soj, Bogoliubov2005,
  Tsilevich:2006}. It serves as a comprehensive review of existing
literature, offering insights into established concepts. While the
content may not be entirely original, its presentation adds value,
particularly from a pedagogical standpoint. Additionally, it provides
a well-curated collection of references for further exploration.

In structuring this discussion, we adopt certain conventions
from~\cite{Wheeler:2010vmq}, while aligning our presentation style
more closely with that of~\cite{Tsilevich:2006}.

@@

\subsection{Phase model}
Consider the \((M+1)\) set of operators \(\{\phi_i,
\phi^\dagger,\mathcal{N}_i\}_{i=0}^M\) such that
\begin{equation}
 [\mathcal{N}_i, \phi_j] = - \phi_i \delta_{i,j} \quad
 [\mathcal{N}_i, \phi_j^\dagger] =  \phi_i^\dagger \delta_{i,j}  \quad 
 [\phi_i, \phi_j^\dagger] =  \pi_i \delta_{i,j}  
\end{equation}
where \(\pi_i =(\ket{0} \bra{0})_i\) is the vacuum projection operator.
These operators can be written as
\begin{equation}
\phi = \sum_{n\geq 0}\ket{n}\bra{n+1}\quad 
\phi^\dagger = \sum_{n\geq 0}\ket{n+1}\bra{n} \quad 
N = \sum_{n\geq 0}n\ket{n}\bra{n}\; ,
\end{equation}
and it is easy to see that \(\phi^\dagger \phi = \bm{1} - \ket{0}
\ket{0}\) and \(\phi\phi^\dagger = \bm{1}\).

@@

These operators appear in the context of quantum optics, and for this
reason, it is referred to as the \emph{phase model}. It corresponds to
the strongly correlated limit of the \(q\)-bosons
model~\cite{Bogoliubov:1997soj}, which we will define shortly.

@@

\subsubsection{Representation}
The representation of the phase model algebra, is constructed using
the vacuum state, defined by \(\ket{0}_i\) by \(\phi_i\ket{0}_i =
0\). In this context, the state with \(n_i\) bosons (oscillators) is
given by \(\ket{n_i}_i = \phi_i^\dagger \ket{0}_i\).

Given the vacuum \(\ket{\bm{0}} = \ket{0}_0\otimes \ket{0}_1
\otimes \cdots \otimes  \ket{0}_M\), the Fock space is defined as 
\begin{equation}
  \mathcal{F} = \bigotimes_{i=0}^M \mathcal{F}_i = 
  \left\{\ket{\vec{n}} = \ket{n_0}\otimes \ket{n_1} \otimes \cdots
  \otimes \ket{n_M} \ | \ n_i \in \mathbb{N} \right\}\; ,
\end{equation}
where the states \(\ket{\vec{n}}\) are defined as 
\begin{equation}
  \ket{\vec{n}} = \ket{n_0}\otimes \ket{n_1} \otimes \cdots \otimes \ket{n_M} 
 \equiv  (\phi_0^\dagger)^{n_0} (\phi_1^\dagger)^{n_1} \cdots  (\phi_M^\dagger)^{n_M} \ket{\bm{0}} \; ,
\end{equation}
with \(\phi_i^\dagger \equiv \bm{1} \otimes \cdots \otimes
\phi_i^\dagger \otimes \cdots \otimes \bm{1}\).  Moreover, it is easy
to see that these states are normalized, that is
\(\bracket{\vec{n}}{\vec{m}}=\delta_{\vec{n}, \vec{m}}\).  Finally,
the actions of the operators \(\mathcal{N}_i\) and \(\pi_i\) are
\begin{equation}
    \pi_i\ket{\vec{n}}  = \delta_{n_i, 0} \ket{\vec{n}} \qquad 
    \mathcal{N}_i\ket{\vec{n}} = n_i \ket{\vec{n}}\; .
\end{equation}

Given a state \(\ket{\vec{n}} = \ket{n_0, n_1, \dots, n_M}\), we
associate a partition \( \lambda = (0^{n_0} 1^{n_1} 2^{n_2} \cdots
M^{n_M})\).  It's worth noting that this correspondence is not unique,
as the partition \(\lambda\) does not account for the number of
particles \(n_0\).  If the total number of particles \(N\) is known,
we can determine \(n_0 = N - \ell(\lambda)\), where \(\ell(\lambda)\)
represents the number of rows in the Young diagram defined by
partition \(\lambda\). This determination is crucial, as in our
subsequent considerations, the \(N\) particle sector remains fixed due
to the integrability of the model. Consequently, the value of \(n_0\)
becomes known once we specify the partition \(\lambda\).

Finally, based on the correspondence we mentioned above,
Wheeler~\cite{Wheeler:2010vmq} defines a map \(\mathcal{M}_\psi:
\mathcal{F} \to \mathcal{F}^{(0)}_\psi\), where
\(\mathcal{F}^{(0)}_\psi\) denotes the Fock space of charged free
fermions constructed from the neutral (fermionic) vacuum.

@@

\subsection{q-Bosons}
The set of operators \(\{b_i, b^\dagger,\mathcal{N}_i\}_{i=0}^M\) constitute
\(M+1\) independent q-boson algebras defined by
\begin{equation}
[\mathcal{N}_i, b_j^\dagger]=\delta_{i,j} b_i^\dagger\; , \quad 
[\mathcal{N}_i, b_j]=-\delta_{i,j}b_i\; , \quad
[b_i, b_j^\dagger]= \delta_{i,j} q^{-2\mathcal{N}_i}  \equiv \delta_{i,j} Q^{\mathcal{N}_i}\; , 
\end{equation}
where we denote the deformation parameter as \(Q = q^{-2}\).

@@

\subsubsection{Representation}
We construct the representation of the q-boson algebra with
modifications of the phase model discussed earlier. Let us define the
\(i\)-th vacuum \(\ket{0}_i\) such that \(b_i\ket{0}_i = 0\). The
representation of this Hilbert space, denoted by \(\mathcal{F}_i^Q\),
is given by \(\ket{n_i}_i \propto b_i^\dagger \ket{0}_i\).

Given the vacuum \(\ket{\bm{0}} = \ket{0}_0\otimes \ket{0}_1
\otimes \cdots \otimes  \ket{0}_M\),
the Fock space is defined as 
\begin{equation}
  \mathcal{F}_{Q} = \bigotimes_{i=0}^M \mathcal{F}_i^Q = 
  \left\{\ket{\vec{n}} = \ket{n_0}\otimes \ket{n_1} \otimes \cdots
  \otimes \ket{n_M} \ | \ n_i \in \mathbb{N} \right\}\; ,
\end{equation}
where the actions of the operators \(\{b_i, b_i^\dagger\}\) are given
by the following relations
\begin{subequations}
\begin{alignat}{3}
    & b_0\ket{n_0} =(1 - \delta_{0, n_0}) \frac{(1 - Q^{n_0})}{(1 - Q)^{1/2}}\ket{n_0 - 1}
    &\hspace{0.5cm}& b_0^\dagger \ket{n_0} =  \frac{1}{(1 - Q)^{1/2}}\ket{n_0 + 1}  \\
    & b_i\ket{n_i} = \frac{(1 - \delta_{0, n_i})}{(1 - Q)^{1/2}}\ket{n_i - 1}
    && b_i^\dagger \ket{n_0} =  \frac{(1 - Q^{n_i+1})}{(1 - Q)^{1/2}}\ket{n_0 + 1}\quad i\neq 0\; .
\end{alignat}
\end{subequations}

@@

This section explores the presence of integrable hierarchies in the
phase model. We demonstrate its correspondence to a Toda hierarchy tau
function and discuss some implications, particularly its connection to
a matrix model. Additionally, we highlight the existence of
correlation functions in this model that satisfy the KP hierarchy
equations. Finally, we argue that these results imply that these
scalar products are also KP hierarchy tau functions, aligning with the
findings of Wheeler~\cite{Wheeler:2010vmq}.

@@

Define two sets of Miwa coordinates \(\bm{t} = (t_1, t_2, \dots)\) and
\(\bm{t}' = (t'_{-1}, t'_{-2}, \dots)\) by
\begin{equation}
  t_p = \frac{1}{p}\sum_{j=1}^N x_j^p\qquad 
  t'_{-p} = \frac{1}{p}\sum_{j=1}^N y_j^p\; .
\end{equation}
One can write the inner product in terms of these coordinates, that is
\begin{equation}
  \mathcal{I}(N, M|\bm{t}, \bm{t}') = \sum_{\lambda\subseteq [N,M]} s_\lambda(\bm{t}) s_\lambda(\bm{t}') \; .
\end{equation}
This expression is known to be a tau function for \(M, N\to \infty\).
As such, utilizing the free fermions representation~\cite{Alexandrov:2012tr}, it can be written as
\begin{equation}
  \lim_{M,N\to \infty}\mathcal{I}(N, M|\bm{t}, \bm{t}')
 = \bra{\bm{0}} e^{\bm{J}_+(\bm{t})} e^{-\bm{J}_-(\bm{t}')} \ket{\bm{0}}\; .
\end{equation}
It is a tau function of the Toda hierarchy with trivial element \(\mathbb{1} \in GL(\infty)\).
It is nothing but the Cauchy's identity
\begin{equation}
  \sum_{\lambda } s_{\lambda}(\bm{t}) s_{\lambda}(\bm{t}')
    = \exp \left( \sum_{m\geq1} m t_m t'_{-m} \right) \; .
\end{equation}

Bringing all these facts together, the truncation for finite \(M\) and
\(N\) also yields tau functions of the Toda hierarchy. More
specifically, according to~\cite{Alexandrov:2012tr, Kharchev:1991gd,
  Zabrodin:2010ii}, the truncation of the tau function corresponds to
the inclusion of a projection operator \(\mathrm{P}^+\) in the
expectation value of the tau function written in the fermionic
representation.

As a final remark, we can also write
\begin{equation}
  H(z,w) = \sum_{k=0}^{M+N} h_k(zw)^{k-1} \; , 
\end{equation}
with \(h_k=1\) if \(k\in [0, M+N-1]\) and \(0\) otherwise. In this
case, one can define a diagonal \((N+M)\times (N+M)\) matrix
\(\mathcal{H} = \textrm{diag}(h_0, \dots, h_{M+N-1})\). Consequently,
if we repeat the arguments above, we find
\begin{equation}
\mathcal{I}(N, M|\bm{t}, \bm{t}') = \sum_{\lambda} h_\lambda s_\lambda(\bm{t}) s_\lambda(\bm{t}') \; .
\end{equation}
Here, \(h_\lambda\) is equal to \(1\) if \(\lambda \in [N,M]\), and it
is zero otherwise.

@@

From the results of~\cite{Zabrodin:2010ii}, see citations therein, we
have an interesting consequence of this representation. Impose the
Bethe equations~(\ref{eq:bethe_eq}) to one set of variables, say \(x_j
= e^{-ip_j} \in \mathbb{S}^1\). Additionally, let us set \(\bm{t}' = -
\bm{t}^\star\). In this particular case, we have
\begin{equation}
\xi(\bm{t}, z) - \xi(\bm{t}', 1/z)  = 2 \ \textrm{Re}\left(\sum_k t_k z^k\right)\; .
\end{equation}
Then, the phase model is equivalent to an ensemble of \(N\) 2D Coulomb
particles on a circle. In this case, we find that the quantities
\(\lambda_{\ell}\) are eigenvalues of a matrix \(U\).

@@

\subsection{Correlation functions}
Bogoliubov has also shown in~\cite{Bogoliubov2005}, that the
correlation functions
\begin{subequations}
\begin{equation}
\begin{split}
\label{eq:correlation}
  A_m(N, M|\bm{x}, \bm{y}\setminus \{y_N\})
  & = \bra{0} \prod_{j=1}^N \mathbb{C}(x_j)
  \prod_{k=1}^{N-1} \mathbb{B}(y_k) \phi_m^\dagger \ket{0}\\
  & =  \prod_{j=1}^N x_{j}^{M/2} \prod_{k=1}^{N-1} y_{j}^{M/2}
  \bra{0} \prod_{j=1}^N C(1/x_j) \prod_{k=1}^{N-1} B(y_k) \phi_m^\dagger \ket{0}
\end{split}
\end{equation}
can be written as
\begin{equation}
  A_m(N, M|\bm{x}, \bm{y}\setminus \{y_N\}) = 
  \frac{(-1)^{N-1}}{y_N^{(N-1)/2}} \prod_{j=1}^N x_{j}^{M/2}
  \prod_{k=1}^{N-1} y_{j}^{M/2}
  \left( \prod_{t<N} \frac{y_N - y_t}{y_t} \right)
  \frac{\det Q}{\det H} \mathcal{I}(N,M|\bm{x}, \bm{y})
\end{equation}
\end{subequations}
where \(Q\) is an \(N\times N\) matrix with components 
\begin{equation}
 Q_{jN} = x_j^{(M + N - 1- 2m)/2} \quad  \textrm{and} \quad 
 Q_{jk} = H_{jk} \; , 
\end{equation}
and \(H_{jk} = H(x_j, y_k)\) are the components of the matrix \(H\)
in~(\ref{eq:h-matrix}).  The components \(Q_{jN}\) are independent of
the coordinates \(y\); therefore, we cannot express the above
expression as a Toda hierarchy tau function.

@@

Furthermore, we define vector field \(\bm{F}(z) = (F_1, \dots, F_N)\),
where the components are given by
\begin{equation}
    F_j (z) = H(z, y_j) \quad \textrm{if} \ j \neq N\; , \qquad \textrm{and}\qquad 
    F_N (z)  = z^{(M + N - 1 - 2m)/2} \; .
\end{equation}
As before, we expand \(F_j(z)\) as the geometric sum
\begin{equation}
  F_j(z) = \frac{1 - (y_j z)^{M + N}}{1 - y_j z} = \sum_{n=0}^{N + M - 1} (y_j z)^n
  \equiv \sum_{n=0}^{M + N -1} f_{j, n} z^n\; .
\end{equation}

With these definitions, we conclude that
\begin{equation}
\mathcal{A}_m(N,M|\bm{x}, \bm{y}) =\frac{\det_{jk} F_j(x_k)}{\Delta(x)}\; .
\end{equation}
From the expansion 
\begin{equation}
  \det_{jk} F_j(x_k) = \det_{jk} \left(  \sum_{n=0}^{M + N -1} f_{j, n} x_k^n \right) \; .
\end{equation}
By utilizing the Cauchy-Binet formula, we can write
\begin{equation}
    \mathcal{A}_m(N,M|\bm{x}, \bm{y})
  = \sum_{0\leq \ell_N\leq \dots \leq \ell_1 \leq N+M}
  \frac{\det_{jk}(f_{j, \ell_k}) \det_{jk}(x_k^{\ell_j})}{\Delta(x)}\; . 
\end{equation}

>>>>

\address{\noindent 
Instituto de Física Teórica, UNESP-Universidade Estadual Paulista, \\
Rua Dr. Bento T. Ferraz 271, Bl. II, São Paulo 01140-070, SP, Brazil
\\ \& \\
Instituto de Física, Universidade de São Paulo,\\
Rua do Matão, Travessa 1371, 05508-090 São Paulo, SP. Brazil
}

@@

\begin{abstract}
This work investigates the intricate relationship between the q-boson
model, a quantum integrable system, and classical integrable systems
such as the Toda and KP hierarchies. Initially, we analyze scalar
products of off-shell Bethe states and explore their connections to
tau functions of integrable hierarchies. Furthermore, we discuss
correlation functions within this formalism, examining their
representations in terms of tau functions, as well as their Schur
polynomial expansions.
\end{abstract}

\date{\today}
\keywords{Integrability, Bethe states, Schur, Hall-Littlewood, Toda, KP}
\subjclass[2020]{82B20, 82B23}

\maketitle

\setcounter{tocdepth}{1}
\tableofcontents

@@

Here, we examine the emergence of classical integrable structures
within correlation functions of the q-boson system. This quantum
integrable system describes q-deformed bosons confined to
a one-dimensional chain~\cite{Bogoliubov:1992, Bogoliubov:1997soj,
  Bogoliubov2005}. Notably, this model bears close relation to the AL
(Ablowitz-Ladik) model, which is an integrable discretization of the
nonlinear Schrödinger equation.

@@

The authors of~\cite{Foda:2008hn, Wheeler:2010vmq} utilize a rich
set of dualities between the ring of symmetric functions and
\(q\)-deformations of the Heisenberg algebra~\cite{Jing1991, Jing1995}
to explore the combinatorial properties of this spin chain. As a
consequence, it has been shown that the scalar product of off-shell
Bethe states are restricted tau functions of the KP
(Kadomtsev-Petviashvili) hierarchy.

@@

This paper builds upon these developments and aims to extend some of
these results. Firstly, we investigate the connections between
the scalar product and tau functions, exploring them in a wider
context, particularly in the case of the Toda
hierarchy~\cite{Takasaki:2018wsv}. It is important to note that both
the AL and the KP hierarchies are reductions of the Toda
hierarchy. This paper takes steps towards resolving this problem,
which is instrumental in understanding the finer details of these
relationships.

Additionally, we explore other correlation functions within this model
and examine how they fit into the classical integrable context. It is
noteworthy that even when the object is not a tau function itself, it
may have an interesting expansion in terms of different types of
Schur polynomials, highlighting some of its combinatorial
properties\footnote{Results involving Schur and Hall-Littlewood
polynomials have been confirmed using the
software~\cite{Araujo:2024piv}.}. We also investigate some of these
expansions in this paper.

In Section 2, we provide a review of the phase model and
q-bosons. This section serves to establish notation and emphasize the
key aspects relevant to our analysis.  Section 3 delves into the
analysis of the phase model and its connections with the KP and Toda
integrable hierarchies. We explore the general expansion of a phase
model correlation function that yields a Toda system solution.  In
Section 4, we shift our focus to the q-bosons. We discuss a
determinantal formula for the scalar product of two off-shell Bethe
states and its expansion in terms of big, supersymmetric Schur
functions, as well as in terms of Kostka-Foulkes polynomials.
Finally, in Section 5, we conclude with an overview of open problems.

@@

This section introduces the \emph{phase} and \emph{\(q\)-boson}
models~\cite{Bogoliubov:1992, Bogoliubov:1997soj, Bogoliubov2005,
  Tsilevich:2006}. It serves as a comprehensive review of existing
literature, offering insights into established concepts.  While the
content does not introduce novel ideas, its presentation adds value,
particularly from a pedagogical standpoint.  In structuring this
discussion, we adopt certain conventions from~\cite{Wheeler:2010vmq},
while aligning our presentation style more closely with that
of~\cite{Tsilevich:2006}.

@@

\subsection{Phase model}
Consider the \((M+1)\) set of operators \(\{\phi_i,
\phi_i^\dagger,\mathcal{N}_i\}_{i=0}^M\) such that
\begin{equation}
 [\mathcal{N}_i, \phi_j] = - \phi_i \delta_{i,j} \quad
 [\mathcal{N}_i, \phi_j^\dagger] =  \phi_i^\dagger \delta_{i,j}  \quad 
 [\phi_i, \phi_j^\dagger] =  \pi_i \delta_{i,j}  
\end{equation}
where \(\pi_i =(\ket{0} \bra{0})_i\) is the vacuum projection operator.
These operators can be written as
\begin{equation}
\phi = \sum_{n\geq 0}\ket{n}\bra{n+1}\quad 
\phi^\dagger = \sum_{n\geq 0}\ket{n+1}\bra{n} \quad 
N = \sum_{n\geq 0}n\ket{n}\bra{n}\; ,
\end{equation}
and it is easy to see that \(\phi^\dagger \phi = \bm{1} - \ket{0}
\ket{0}\) and \(\phi\phi^\dagger = \bm{1}\).

@@

These operators appear in the context of quantum optics, and for this
reason, this model is referred to as the \emph{phase model}. It corresponds to
the strongly correlated limit of the \(q\)-bosons
model~\cite{Bogoliubov:1997soj}, which we will define shortly.

@@

\subsubsection{Representation}
The representation of the phase model algebra is constructed using
the vacuum state defined by \(\ket{0}_i\) by \(\phi_i\ket{0}_i =
0\). In this context, the state with \(n_i\) bosons (oscillators) is
given by \(\ket{n_i}_i = (\phi_i^\dagger)^{n_i} \ket{0}_i\).

Given the vacuum \(\ket{\bm{0}} = \ket{0}_0\otimes \ket{0}_1
\otimes \cdots \otimes  \ket{0}_M\), the Fock space is defined as 
\begin{equation}
  \mathcal{F} = \bigotimes_{i=0}^M \mathcal{F}_i = \mathrm{span}
  \left\{\ket{\vec{n}} = \ket{n_0}\otimes \ket{n_1} \otimes \cdots
  \otimes \ket{n_M} \ | \ n_i \in \mathbb{N} \right\}\; ,
\end{equation}
where the states \(\ket{\vec{n}}\) are defined as 
\begin{equation}
  \ket{\vec{n}} = \ket{n_0}\otimes \ket{n_1} \otimes \cdots \otimes \ket{n_M} 
 \equiv  (\phi_0^\dagger)^{n_0} (\phi_1^\dagger)^{n_1} \cdots  (\phi_M^\dagger)^{n_M} \ket{\bm{0}} \; ,
\end{equation}
with \(\phi_i^\dagger \equiv \bm{1} \otimes \cdots \otimes
\phi_i^\dagger \otimes \cdots \otimes \bm{1}\).  Moreover, it is easy
to see that these states are normalized, that is
\(\bracket{\vec{n}}{\vec{m}}=\delta_{\vec{n}, \vec{m}}\).  Finally,
the actions of the operators \(\mathcal{N}_i\) and \(\pi_i\) are
\begin{equation}
    \pi_i\ket{\vec{n}}  = \delta_{n_i, 0} \ket{\vec{n}} \qquad 
    \mathcal{N}_i\ket{\vec{n}} = n_i \ket{\vec{n}}\; .
\end{equation}

Given a state \(\ket{\vec{n}} = \ket{n_0, n_1, \dots, n_M}\), we
associate a partition \( \lambda = (1^{n_1} 2^{n_2} \cdots
M^{n_M})\).  It's worth noting that this correspondence is not unique,
as the partition \(\lambda\) does not account for the number of
particles \(n_0\).  If the total number of particles \(N\) is known,
we can determine \(n_0 = N - \ell(\lambda)\), where \(\ell(\lambda)\)
represents the number of rows in the Young diagram defined by
partition \(\lambda\).  This aspect is crucial because, in our
subsequent considerations, the \(N\) particle sector remains fixed
owing to the integrability of the model. Consequently, the value of
\(n_0\) becomes known once we specify the partition \(\lambda\).

Finally, based on the correspondence we mentioned in the introduction,
Wheeler~\cite{Wheeler:2010vmq} defines a map \(\mathcal{M}_\psi:
\mathcal{F} \to \mathcal{F}^{(0)}_\psi\), where
\(\mathcal{F}^{(0)}_\psi\) denotes the Fock space of charged free
fermions constructed from the neutral (fermionic) vacuum.

@@

\subsection{q-Bosons}
The set of operators \(\{b_i, b_i^\dagger,\mathcal{N}_i\}_{i=0}^M\) constitute
\(M+1\) independent q-boson algebras defined by
\begin{equation}
[\mathcal{N}_i, b_j^\dagger]=\delta_{i,j} b_i^\dagger\; , \quad 
[\mathcal{N}_i, b_j]=-\delta_{i,j}b_i\; , \quad
[b_i, b_j^\dagger]= \delta_{i,j} q^{-2\mathcal{N}_i}  \equiv \delta_{i,j} Q^{\mathcal{N}_i}\; , 
\end{equation}
where we denote the deformation parameter as \(Q = q^{-2}\).

@@

\subsubsection{Representation}
Let us define the \(i\)-th vacuum \(\ket{0}_i\) such that
\(b_i\ket{0}_i = 0\). The representation of this Hilbert space,
denoted by \(\mathcal{F}_i^Q\), is defined by the states
\(\ket{n_i}_i \propto (b_i^\dagger)^{n_i} \ket{0}_i\).

Given the vacuum \(\ket{\bm{0}} = \ket{0}_0\otimes \ket{0}_1
\otimes \cdots \otimes  \ket{0}_M\),
the Fock space is defined as 
\begin{equation}
  \mathcal{F}_{Q} = \bigotimes_{i=0}^M \mathcal{F}_i^Q = 
  \mathrm{span}\left\{\ket{\vec{n}} = \ket{n_0}\otimes \ket{n_1} \otimes \cdots
  \otimes \ket{n_M} \ | \ n_i \in \mathbb{N} \right\}\; ,
\end{equation}
where the actions of the operators \(\{b_i, b_i^\dagger\}\) are given
by the following relations
\begin{subequations}
\begin{alignat}{3}
    & b_0\ket{n_0} =(1 - \delta_{0, n_0}) \frac{(1 - Q^{n_0})}{(1 - Q)^{1/2}}\ket{n_0 - 1}
    &\hspace{0.5cm}& b_0^\dagger \ket{n_0} =  \frac{1}{(1 - Q)^{1/2}}\ket{n_0 + 1}  \\
    & b_i\ket{n_i} = \frac{(1 - \delta_{0, n_i})}{(1 - Q)^{1/2}}\ket{n_i - 1}
    && b_i^\dagger \ket{n_0} =  \frac{(1 - Q^{n_i+1})}{(1 - Q)^{1/2}}\ket{n_0 + 1}\quad i\neq 0\; .
\end{alignat}
\end{subequations}

@@

This section explores the presence of integrable hierarchies in the
phase model. We demonstrate its relations with the Toda hierarchy tau
function and discuss some implications, particularly its connection to
a matrix model. We also argue that these results imply that the scalar
products in the model also serve as KP hierarchy tau functions,
consistent with the findings of
Wheeler~\cite{Wheeler:2010vmq}. Additionally, we highlight the
existence of correlation functions in this model that satisfy the KP
hierarchy equations.

@@

Define two sets of Miwa coordinates \(\bm{t} = (t_1, t_2, \dots)\) and
\(\bm{t}' = (t'_{-1}, t'_{-2}, \dots)\) by
\begin{equation}
  t_p = \frac{1}{p}\sum_{j=1}^N x_j^p\qquad 
  t'_{-p} = \frac{1}{p}\sum_{j=1}^N y_j^p\; .
\end{equation}
One can write the inner product in terms of these coordinates, that is
\begin{equation}
  \mathcal{I}(N, M|\bm{t}, \bm{t}') = \sum_{\lambda\subseteq [N,M]} s_\lambda(\bm{t}) s_\lambda(\bm{t}') \; .
\end{equation}
This expression is known to be a tau function for \(M, N\to \infty\).
As such, utilizing the free fermions representation~\cite{Alexandrov:2012tr}, it can be written as
\begin{equation}
  \lim_{M,N\to \infty}\mathcal{I}(N, M|\bm{t}, \bm{t}')
 = \bra{\bm{0}} e^{\bm{J}_+(\bm{t})} e^{-\bm{J}_-(\bm{t}')} \ket{\bm{0}}\; .
\end{equation}
It is a tau function of the Toda hierarchy with trivial element
\(\mathbb{1} \in GL(\infty)\), and it is nothing but the Cauchy's
identity
\begin{equation}
  \sum_{\lambda } s_{\lambda}(\bm{t}) s_{\lambda}(\bm{t}')
    = \exp \left( \sum_{m\geq1} m t_m t'_{-m} \right) \; .
\end{equation}

Bringing all these facts together, the truncation for finite \(M\) and
\(N\) also yields tau functions of the Toda hierarchy. More
specifically, according to~\cite{Alexandrov:2012tr, Kharchev:1991gd,
  Zabrodin:2010ii}, the truncation of the tau function corresponds to
the inclusion of a projection operator in the expectation value of the
tau function written in the fermionic representation.

As a final remark, we can also write
\begin{equation}
  H(z,w) = \sum_{k=0}^{M+N-1} h_k(zw)^{k} \; , 
\end{equation}
with \(h_k=1\) if \(k\in [0, M+N-1]\) and \(0\) otherwise. In this
case, one can define a diagonal \((N+M)\times (N+M)\) matrix
\(\mathcal{H} = \textrm{diag}(h_0, \dots, h_{M+N-1})\). Consequently,
if we repeat the arguments above, we find
\begin{equation}
\mathcal{I}(N, M|\bm{t}, \bm{t}') = \sum_{\lambda \subseteq [N,M]} h_\lambda s_\lambda(\bm{t}) s_\lambda(\bm{t}') \; .
\end{equation}
Here, \(h_\lambda\) is equal to \(1\) if \(\lambda \in [N,M]\), and it
is zero otherwise.

@@

From the results of~\cite{Zabrodin:2010ii}, see citations therein, we
have an interesting consequence of this representation. Impose the
Bethe equations~(\ref{eq:bethe_eq}) to one set of variables, say \(x_j
= e^{-ip_j} \in \mathbb{S}^1\). Additionally, let us set \(\bm{t}' = -
\bm{t}^\star\). In this particular case, we have
\begin{equation}
\xi(\bm{t}, z) - \xi(\bm{t}', 1/z)  = 2 \ \textrm{Re}\left(\sum_k t_k z^k\right)\; .
\end{equation}
Then, the phase model is equivalent to an ensemble of \(N\) 2D Coulomb
particles on a circle. In this case, we find that the quantities
\(z_{\ell}\) are eigenvalues of a matrix \(U\).

@@

\subsection{Correlation functions}
Bogoliubov has also shown, in~\cite{Bogoliubov2005}, that the
correlation functions
\begin{subequations}
\begin{equation}
\begin{split}
\label{eq:correlation}
  A_m(N, M|\bm{x}, \bm{y}\setminus \{y_N\})
  & = \bra{0} \prod_{j=1}^N \mathbb{C}(x_j)
  \prod_{k=1}^{N-1} \mathbb{B}(y_k) \phi_m^\dagger \ket{0}\\
  & =  \prod_{j=1}^N x_{j}^{M/2} \prod_{k=1}^{N-1} y_{j}^{M/2}
  \bra{0} \prod_{j=1}^N C(1/x_j) \prod_{k=1}^{N-1} B(y_k) \phi_m^\dagger \ket{0}
\end{split}
\end{equation}
can be written as
\begin{equation}
  A_m(N, M|\bm{x}, \bm{y}\setminus \{y_N\}) = 
  \frac{(-1)^{N-1}}{y_N^{(N-1)/2}} \prod_{j=1}^N x_{j}^{M/2}
  \prod_{k=1}^{N-1} y_{j}^{M/2}
  \left( \prod_{t<N} \frac{y_N - y_t}{y_t} \right)
  \frac{\det Q}{\det H} \mathcal{I}(N,M|\bm{x}, \bm{y})
\end{equation}
\end{subequations}
where \(Q\) is an \(N\times N\) matrix with components 
\begin{equation}
 Q_{jN} = x_j^{(M + N - 1- 2m)/2} \quad  \textrm{and} \quad 
 Q_{jk} = H_{jk} \; , 
\end{equation}
and \(H_{jk} = H(x_j, y_k)\) are the components of the matrix \(H\)
in~(\ref{eq:h-matrix}).  The components \(Q_{jN}\) are independent of
the coordinates \(y\); therefore, we cannot express the above
expression as a Toda hierarchy tau function.

@@

Furthermore, we define vector field \(\bm{F}(z) = (F_1, \dots, F_N)\),
where its components are given by
\begin{equation}
    F_j (z) = H(z, y_j) \quad \textrm{if} \ j \neq N\; , \qquad \textrm{and}\qquad 
    F_N (z)  = z^{(M + N - 1 - 2m)/2} \; .
\end{equation}
As before, we expand \(F_j(z)\), \(j\neq N\), as the geometric sum
\begin{equation}
  F_j(z) = \frac{1 - (y_j z)^{M + N}}{1 - y_j z} = \sum_{n=0}^{N + M - 1} (y_j z)^n
  \equiv \sum_{n=0}^{M + N -1} f_{j, n} z^n \; , \quad  f_{j, n} = y_j^n\; .
\end{equation}
We can also express the function \(F_N(z)\) in this form by setting
\(f_{N,n} = \delta_{n,(M - N - 1 - 2m)/2}\) and requiring \((M - N -
1)\) to be an even integer.

With these definitions, we conclude that
\begin{equation}
\mathcal{A}_m(N,M|\bm{x}, \bm{y}) =\frac{\det_{jk} F_j(x_k)}{\Delta(x)}\; .
\end{equation}
From the expansion 
\begin{equation}
  \det_{jk} F_j(x_k) = \det_{jk} \left(  \sum_{n=0}^{M + N -1} f_{j, n} x_k^n \right) \; ,
\end{equation}
and utilizing the Cauchy-Binet formula, we get
\begin{equation}
    \mathcal{A}_m(N,M|\bm{x}, \bm{y})
  = \sum_{0\leq \ell_N\leq \dots \leq \ell_1 \leq N+M}
  \frac{\det_{jk}(f_{j, \ell_k}) \det_{jk}(x_k^{\ell_j})}{\Delta(x)}\; . 
\end{equation}
